### 2017-07-03 paper/introduction.tex
OK

\noindent{\em Probabilistic programs with nondeterminism.} 
Randomness play a fundamental role in many areas across science, and in 
computer science in particular.
In applications such as stochastic network protocols~\cite{BaierBook,prism},
randomized algorithms~\cite{RandBook,RandBook2}, machine learning~\cite{},
NATURE PAPER, the probabilistic behavior must be considered to faithfully model
the underlying dynamic system.
The extension of classical imperative programs with \emph{random value generators}, 
that produce random values according to some desired probability distribution, 
naturally gives probabilistic programs.
Along with probability, nondeterminism also plays a crucial role.
In particular in program analysis, for effective analysis of large programs,
all variables cannot be considered, and abstraction ignores some variables,
and the worst-case analysis is represented by adversarial nondeterminism.
Hence probabilistic programs with nondeterminism has become an active and 
important research area in program analysis.

\smallskip\noindent{\em Termination problem.} 
In static analysis of programs the most basic, as well most important, 
liveness property is the {\em termination} problem.
While for non-probabilistic programs the termination question asks whether
an input program {\em always} terminates, for probabilistic programs 
the termination questions accounts for the probabilistic behaviors. 
The most basic and fundamental extensions of the termination problem 
for probabilistic programs are:

@@

\smallskip\noindent{\em Ranking functions and ranking supermartingales (RSMs).}
The key technique that applies for liveness analysis of non-probabilistic programs is 
the notion of {\em ranking functions}, which provides a sound and complete 
method for termination of non-probabilistic programs~\cite{rwfloyd1967programs}.
There exist a wide variety of different approaches for construction 
of ranking functions for non-probabilistic programs~\cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/ColonS01,DBLP:conf/vmcai/PodelskiR04,DBLP:conf/pods/SohnG91}.
The generalization of ranking function to probabilistic programs is achieved through the
{\em ranking supermartingales (RSMs)}~\cite{}.
The ranking supermartingales provide a powerful and automated approach for termination 
analysis of probabilistic programs~\cite{SRIRAM,HOLGER,HONGFEI}, and algorithmic 
approaches for special cases such as linear and polynomial RSMs has
also been considered~\cite{POPL,CAV,POPL}.

\smallskip\noindent{\em .} 
While a impressive set of theoretical results related to RSMs has been
established, for probabilistic programs with nondeterminism the current approaches  
are only applicable to academic examples of variants of random walks. 
The key reason can be understood as follows: even for non-probabilistic programs while 
ranking functions are sound and complete, they do not necessarily provide a practical 
approach. 
In contrast, as a practical and scalable approach for non-probabilistic programs the 
notion of lexicographic ranking functions has been widely studied~\cite{}. 
Algorithmic approaches for lexicographic ranking functions allow the termination analysis 
to be applicable to real-world non-probabilistic programs (after abstraction).
However both the theoretical foundations as well as practical approaches related to 
such lexicographic ranking functions are completely missing for probabilistic programs,
which we address in this work.

\smallskip\noindent{\em Our contributions.} 
In this work our main contributions ranges from theoretical foundations of lexicographic
RSMs, to algorithmic approaches for them, to experimental results
showing their applicability for programs.
We describe our main contributions below:
\begin{enumerate}

\item {\em Theoretical foundations.} 
First, we introduce the notion of lexicographic RSMs, and show that
such supermartingales ensures almost-sure termination (Theorem~\ref{} in Section~\ref{}). 
Our first result is a purely mathematical result that introduces a new concept, and 
proves almost-sure termination, that is independent of any probabilistic program.
Based on the mathematical result we show that probabilistic programs with nondeterminism
the existence of a lexicographic ranking functions with respect to an invariant ensures
almost-sure termination (Theorem~\ref{} in Section~\ref{}).

@@

\item {\em Asymptotic bounds.} 
In general, the existence of lexicographic RSMs does not imply positive
termination. 
In other words, we present an example (Example~\ref{}) where lexicographic RSM exists ensuring
almost-sure termination, but the expected termination time is infinite.
We then present a natural restriction (i.e., a subclass of programs) such the lexicographic 
RSMs not only imply positive termination, but even asymptotic bounds on the expected termination 
time can be derived from them (see Theorem~\ref{}).

>>>>

\noindent{\em Probabilistic programs with nondeterminism.} 
Randomness play a fundamental role in many areas across science, and in 
computer science in particular.
In applications such as stochastic network protocols~\cite{BaierBook,prism},
randomized algorithms~\cite{RandBook,RandBook2}, machine learning~\cite{},
NATURE PAPER, the probabilistic behavior must be considered to faithfully model
the underlying dynamic system.
The extension of classical imperative programs with \emph{random value generators}, 
that produce random values according to some desired probability distribution, 
naturally gives rise probabilistic programs.
Along with probability, nondeterminism also plays a crucial role.
In particular in program analysis, for effective analysis of large programs,
all variables cannot be considered, and abstraction ignores some variables,
and the worst-case analysis is represented by adversarial nondeterminism.
Hence probabilistic programs with nondeterminism has become an active and 
important research area in program analysis.

\smallskip\noindent{\em Termination problem.} 
In static analysis of programs the most basic, as well most important, 
liveness property is the {\em termination} problem.
While for non-probabilistic programs the termination question asks whether
an input program {\em always} terminates, for probabilistic programs 
the termination questions must account for the probabilistic behaviors. 
The most basic and fundamental extensions of the termination problem 
for probabilistic programs are:

@@

\smallskip\noindent{\em Ranking functions and ranking supermartingales (RSMs).}
The key technique that applies for liveness analysis of non-probabilistic programs is 
the notion of {\em ranking functions}, which provides a sound and complete 
method for termination of non-probabilistic programs~\cite{rwfloyd1967programs}.
There exist a wide variety of approaches for construction 
of ranking functions for non-probabilistic programs~\cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/ColonS01,DBLP:conf/vmcai/PodelskiR04,DBLP:conf/pods/SohnG91}.
The generalization of ranking functions to probabilistic programs is achieved through the
{\em ranking supermartingales (RSMs)}~\cite{}.
The ranking supermartingales provide a powerful and automated approach for termination 
analysis of probabilistic programs~\cite{SRIRAM,HOLGER,HONGFEI}, and algorithmic 
approaches for special cases such as linear and polynomial RSMs has
also been considered~\cite{POPL,CAV,POPL}.

\smallskip\noindent{\em Practical limitations of existing approaches.} 
While a impressive set of theoretical results related to RSMs has been
established~\cite{}, for probabilistic programs with nondeterminism the current approaches  
are only applicable to academic examples of variants of random walks. 
The key reason can be understood as follows: even for non-probabilistic programs while 
ranking functions are sound and complete, they do not necessarily provide a practical 
approach. 
In contrast, as a practical and scalable approach for non-probabilistic programs the 
notion of lexicographic ranking functions has been widely studied~\cite{}. 
Algorithmic approaches for lexicographic ranking functions allow the termination analysis 
to be applicable to real-world non-probabilistic programs (after abstraction).
However both the theoretical foundations as well as practical approaches related to 
such lexicographic ranking functions are completely missing for probabilistic programs,
which we address in this work.

\smallskip\noindent{\em Our contributions.} 
In this work our main contributions range from theoretical foundations of lexicographic
RSMs, to algorithmic approaches for them, to experimental results
showing their applicability for programs.
We describe our main contributions below:
\begin{enumerate}

\item {\em Theoretical foundations.} 
First, we introduce the notion of lexicographic RSMs, and show that
such supermartingales ensures almost-sure termination (Theorem~\ref{} in Section~\ref{}). 
Our first result is a purely mathematical result that introduces a new concept, and 
proves almost-sure termination, that is independent of any probabilistic program.
Based on the mathematical result we show that probabilistic programs with nondeterminism
the existence of a lexicographic RSM with respect to an invariant ensures
almost-sure termination (Theorem~\ref{} in Section~\ref{}).

@@

\item {\em Asymptotic bounds.} 
In general, the existence of lexicographic RSMs does not imply positive
termination. 
In other words, we present an example (Example~\ref{}) where a lexicographic RSM exists ensuring
almost-sure termination, yet the expected termination time is infinite.
We then present a natural restriction (i.e., a subclass of programs) such that the lexicographic 
RSMs not only imply positive termination, but even asymptotic bounds on the expected termination 
time can be derived from them (see Theorem~\ref{}).
